### 2016-10-22 paper/paper.tex
PAPER: initial template

\documentclass{vldb}

\usepackage{graphicx}

\title{Title goes here}
\author{Filippo Sestini \\ Vrije Universiteit \\ email@host.com \and
  Davide Dal Bianco \\ Vrije Universiteit \\ email@host.com}

\begin{document}
\maketitle

\begin{abstract}
Abstract goes here.  
\end{abstract}

\section{Introduction}
% what is this project about and why is it interesting?

\section{Related Work}
% meaningfully summarize related literature to explain what this project builds
% on, in particular, discuss the two research papers chosen with this project
% (and discussed in your presentation)

\section{Research Questions}
% which questions is this project trying to answer, and/or hypotheses to
% investigate? These should cover both the project topic, as well as the
% technological side (aptness of the tools for the job, scalability of the
% solution).

\section{Project Setup}
% the steps taken during the project.

\section{Experiments}
% a description of the experiments and their results.

\section{Conclusion}
% revisit the research questions and hypotheses and try to answer them. Any new
% questions? Insights in the usability of the employed technology for particular
% tasks?

\end{document}

### 2016-10-22 paper/paper.tex
PAPER: add some subsections
PAPER: add even more subsections
Added subsection opensky

@@
 
+\subsection{Bringing up OpenSky}
+Most of aircraft are provided with ADB-S transmitter, which send continuosly information about the status and the position of the aircraft to guarantee better air traffic control.
+The messages are not encripted end, in most region, this technology will become mandatory in 2020.
+For these reasons OpenSky network was recently born.
+\\
+OpenSky is an open source network that aggregates ADS-B messages collected by volunteers around the globe.
+The dataset is accessible to everyone for non-profit purposes and the data is provided in the raw format, in order to perform research and statistics about air traffic.
+\\
+However there are some limitation to the data.
+Only part of Europe has a good coverage, which means there are many areas where is not yet possible to detect airplanes informations.
+Furthermore messages are collected by volunteers and there is not guarantee that antennas are working properly 24/7.
+
+    \subsubsection{ADS-B messages}
+
+    The ADS-B standard defines different kinds of messages which are sent at different frequencies:
+    \begin{itemize}
+        \item \textit{Aircraft identification} messages contain the callsign of the airplane;
+        \item \textit{Airborne position} messages contain altitude, latititude and longitude of the airplane and are sent twice per second;
+        \item \textit{Airborne velocity} messages contain climb rate and velocity of the airplane.
+    \end{itemize}
+    In addition to these messages, other emergencies, priority, capability, navigation accuracy category, and operational modes messages could be broadcasted.
+
+\subsection{OpenSky Network}
+
+\subsection{Route detection}
+
 \section{Research Questions}
@@
 
+\begin{itemize}
+  \item What are the standard routes?
+  \item Which flights and airlines diverge from them? In which airports?
+  \item How much CO2 would be saved if aircrafts followed a straight-line route
+  instead of the standard one?
+\end{itemize}
+
+Additionally
+
+\begin{itemize}
+  \item How to use raw ADS-B data to identify flights?
+\end{itemize}
+
 \section{Project Setup}
 % the steps taken during the project.
+
+\subsection{Technologies}
+% spark, scala, etc
+% python, haskell for scripting, etc
+
+\subsection{Algorithms}
+
+\subsubsection{Positions}
+
+\subsubsection{Flights}
+
+\subsubsection{Routes}
+
+\subsection{Deployment on the hadoop cluster}
+% how we actually implemented the algorithms on the hadoop cluster
 

### 2016-10-22 paper/paper.tex
Added subsection route clustering

@@
 
-\subsection{Bringing up OpenSky}
+\subsection{OpenSky Network}
 Most of aircraft are provided with ADB-S transmitter, which send continuosly information about the status and the position of the aircraft to guarantee better air traffic control.
@@
 
-\subsection{OpenSky Network}
-
 \subsection{Route detection}
+A way to reduce the number of air routes is illustrated in \textit{J.
+DeArmon et al, Air Route Clustering for a Queuing Network Model of the National Airspace System}.
+The research showed that similar routes can be grouped to reduce the number of path between a pair of airports thus decrease air traffic congestion.
+The same approach could be used to detect standard routes where aircraft fly, given the departure and arrival airports.
+The first prototype of this model has been developed to improve the air traffic flow in the National Airspace System of the USA.
 

### 2016-10-22 visualization/public/airlines.js
Added airlines table

@@ -1,0 +1,1 @@
+var airlines = [{"name":"Aerodienst GmbH","nDiverging":1},{"name":"Airbus Industrie","nDiverging":3},{"name":"Air Berlin","nDiverging":16},{"name":"Austrian Airlines","nDiverging":1},{"name":"Alitalia","nDiverging":23},{"name":"Air Transat","nDiverging":1},{"name":"British Airways","nDiverging":2},{"name":"Croatia Airlines","nDiverging":2},{"name":"Czech Airlines","nDiverging":6},{"name":"Ethiopian Airlines","nDiverging":1},{"name":"European Air Transport","nDiverging":22},{"name":"Eurowings","nDiverging":1},{"name":"easyJet","nDiverging":15},{"name":"Federal Express","nDiverging":2},{"name":"Germanwings","nDiverging":20},{"name":"KLM Royal Dutch Airlines","nDiverging":13},{"name":"Lufthansa","nDiverging":42},{"name":"Lufthansa Cargo","nDiverging":1},{"name":"Luxair","nDiverging":3},{"name":"MNG Airlines","nDiverging":1},{"name":"Neos","nDiverging":1},{"name":"Qatar Airways","nDiverging":1},{"name":"Ryanair","nDiverging":24},{"name":"Swiss International Air Lines","nDiverging":14},{"name":"Swissair","nDiverging":14},{"name":"Stuttgarter Flugdienst","nDiverging":1},{"name":"Thomsonfly","nDiverging":1},{"name":"Tiphook PLC","nDiverging":2},{"name":"United Parcel Service","nDiverging":1},{"name":"Vistajet","nDiverging":1},{"name":"Vueling Airlines","nDiverging":9},{"name":"Wizz Air","nDiverging":6}]

### 2016-10-22 visualization/public/diverging_airlines.js

@@ -1,1 +1,0 @@
-var diverging_airlines = [{"name":"Airbus Industrie","nDiverging":1},{"name":"Miramichi Air Service","nDiverging":1},{"name":"Stuttgarter Flugdienst","nDiverging":1},{"name":"Vistajet","nDiverging":3}]

### 2016-10-22 paper/paper.tex
Add content to Project Setup section

@@
 
+Decoding the position of an aircraft from ADS-B position messages is more
+complicated than one would expect. Information on latitude and longitude is
+divided into two types of position messages: odd and even. Both these types of
+messages are needed to correctly decode the position. This is according to a
+format called Compact Position Reporting (CPR), which allows to encode higher
+resolution information in less bits, but is obviously less straight-forward to
+decode.
+
+For the actual decoding in our algorithm we used the java-adsb library, which
+however makes the effective parallelization of the entire operation more
+challenging. In partocular, the decoding procedure implemented by the library is
+essentially sequential, in that all position messages of a particular aircraft
+must be submitted in order of increasing timestamp. This renders the
+distribution and parallelization of the process practically impossible.
+
+However, even though the positions for a single aircraft must be decoded
+sequentially by a single node, it is still possible to distribute the data by
+aircraft, allowing us to spread the work among the cluster. The distribution
+involves grouping operations that are known to be particularly expensive, since
+they require a lot of data to be shuffled and shared between the nodes.
+Nevertheless, we did not see alternative approaches to the problem, given the
+(quite intrinsic) limitations of the decoding process.
+
+The position decoding algorithm goes through the following phases:
+
+\begin{enumerate}
+  \item Extract the raw messages and their timestamp from the avro-encoded data
+  \item Decode the messages from the raw representation
+  \item Filter out messages that are not position messages
+  \item Order the messages according to their timestamp
+  \item Group the messages by ICAO
+  \item For each ICAO, go through the messages in order of timestamp and decode
+  the positions
+\end{enumerate}
+
+Notice that, even if we were able to distribute the work by ICAO, there is still
+a very high number of messages to be processed by a single worker node. In order
+to avoid memory problems, we decided to split the decoding process between days,
+therefore running the above algorithm separately for each day that we had
+available in the dataset. This choice has obvious consequences on the phases
+that follow, as we were not able to detect flights between two different days.
+From our observations, however, we concluded that these flights are not very
+frequent, so at the end of the day the final results did not suffer too much.
+
+% Format of the output?
+
 \subsubsection{Flights}
 
+\paragraph{Identification of discrete flights}
+
+Having extracted all position information from the raw avro files, there is now
+the problem of identifying discrete flights from it.
+Given a sequence of positions for a given aircraft, on a particular day, we
+though about how to group this information in chunks, each chunk representing a
+potential flight. An important insight is given by observing the data: ADS-B
+messages systematically stop being received by the sensors when the aircraft
+approaches the ground. It follows that there must be a pause in the data between
+two different flights, and in particular between the landing of a flight and the
+take-off of another.
+
+Another point to consider is the characteristics of position messages received
+during a flight. Position messages are sent by the aircrafts several times per
+minute, so two subsequent entries are likely to differ very slightly in their
+space, altitude and timestamp values. If they are not, then these messages are
+not considered good enough for our purposes.
+
+Position points for a given aircraft are thus grouped together according to a
+simple clustering algorithm: all points are analyzed in order of timestamp, and
+two subsequent points are considered belonging in the same cluster if and only
+if they are no more than 20 minutes apart, their distance is no more than 20
+kilometers and their altitudes differ no more than 200 meters.
+
+Each group is then further analyzed to determine if it actually corresponds to a
+single flight or not. Again, the analysis is fairly simple, and it considers
+altitude information of the points to detect take-offs and landings. In
+particular, a group of points is considered a valid flight if and only if it
+starts and ends with points at an altitude of 3000 km or below, and shows an
+ascent as well as a descent. Ascents and descents are defined as sequences of
+points which altitudes rise, respectively decrease, steadily for 2000 meters.
+
+\paragraph{Departure and arrival airports}
+
+A flight is given by a sequence of positions, as well as departure and arrival
+airports. There two last informations, however, are not provided by the data and
+must be inferred in some way. Our approach was to consider the first and last
+position point of a flight, and determine the airports as the closest ones to
+these points, in terms of euclidean distance.
+
+To detect the airports, we used the help of an external dataset containing a
+list of airports together with their locations. This dataset originally
+contained much more data than we needed, as it also listed small non-commercial
+airports. In order for our algorithm to be reasonably correct, we had to trim
+the dataset first removing the airports without a IATA code, and then keeping
+only commercial airports with the help of an external list we found on the
+Internet. This filtering reduced the list from XXX to XXX entries.
+The resulting number of entries is still a little bit high for our purposes, but
+it was nevertheless the best dataset we could get. The alternative would have
+been to build a dataset by hand, which seemed extremely time consuming.
+% TODO ^^^^^^^^^^
+
+We specifically set the threshold of 3000 m in the flight detection algorithm
+for purpose of airport detection: after some observations of read flights on the
+website FlightRadar24, we concluded that when an aircraft goes below an altitude
+of 3000 m, its position is very close to the take-off/landing airport. The
+airport identification works reasonably well in this setting, and with the
+external dataset. It is of couse not perfect, for two main reasons:
+
+\begin{itemize}
+  \item The dataset may still contain minor airports that are very close to
+  bigger commercial ones that are the real target of a flight; in these
+  situations, it is possible that the minor airport gets erroneously selected;
+  \item A threshold of 3000 m may noy be sufficient in areas that contain many
+  big commercial airports that are very close to each other; one example of such
+  area is London.
+\end{itemize}
+
+Having considered these issues, we still settled for the simplest solution, as
+we did not find alternative ways to detect airports given the data available.
+
+\paragraph{Sectors}
+
+The route detection algorithm, as explained in details below, uses a
+representation of flights and routes as a sequence of sectors, where each sector
+represents portion of the world map. For this reason, the flights detection
+phase outputs flights as sequences of sectors rather than position points. This
+has two main advantages:
+
+\begin{itemize}
+  \item It prepares the data to be easily processed by the following phase of
+  the algorithm
+  \item It represents flights with much less space than as a highly-dense
+  sequence of points
+\end{itemize}
+
+We could have considered alternative ways to eliminate redundant information and
+reduce the amount of data needed to represent a single flight, but the routes
+detection algorithm would stil have transformed those representations to
+sequences of sectors. It follows that converting directly from position points
+to sectors was obviously the best approach.
+
+In practice, a sequence of position points is translated in a sequence of
+sectors by simply truncating the latitude and longitude part of a point to one
+decimal digit. Strings of values with the same truncated latitude and
+longitude represent the same sector, and are compressent into one. This process
+yields sectors of about 10 km by 10 km, which is a good approximation for our
+purposes, where sub-kilometer precision is not needed.
+
 \subsubsection{Routes}
+
+The most important part of the process is the detection of routes between pairs
+of airports. Standard routes between a pair airports are likely to be used by
+the majority of flights between those airports. Our algorithm relies on the
+assumption that the converse should also be true, namely that if a route is used
+by the majority of flights going from an airport to another, than that route
+must represent a (possibly, the) standard route between that pair of airports.
+
+The actual route detection process relies on an aggregative clustering algorithm
+that groups together flights according to a distance metric. The results of the
+clustering are used to determine the standard route between a certain pair of
+airports. More precisely, the algorithm goes through the following steps for
+each pair of airports:
+
+\begin{enumerate}
+  \item Apply the clustering algorithm to all the flights that have the
+  current pair of airports as endpoints.
+  \item Select the cluster that classifies the highest number of routes.
+  \item Select a route that constitues the representative of
+  that cluster as the one having the least distance between al the other
+  routes in the cluster
+  \item The representative of this cluster determines the standard route
+  between the pair of airports
+  \item All the routes classified by the other clusters represent routes that
+  diverge from the standard route
+\end{enumerate}
+
+The approach described above has the advantage of being simple and generally
+correct in most situations. It has, of course, some problems, that come in
+particular from its being fairly naive. First of all, it goes under the
+assumption that the standard route, officially prescribed by the
+appropriate institutions, is the one that is used by most flights. This
+assumption may be invalidated in two obvious ways, among possibly others:
+
+\begin{itemize}
+  \item The analysis clearly does not take into account flights, and therefore
+  routes that cannot make through the route detection phase of the algorithm.
+  Our dataset is not extremely reliable, so say the least. The OpenSky Network
+  has, and the time or writing this paper, a fairly good coverage of western
+  Europe, but some sports may still be poorly detected by the sensors. If a
+  standard route happens to pass through an area that is poorly covered, it is
+  likely that other, non standard routes will be selected by the algorithm as
+  standard, just because they can be detected.
+  \item There may be reasons or events, such as military conflicts, for which
+  airplanes are forced to systematically deviate from a standard route, for a
+  period of time that can span days or even months. These temporary anomalies
+  clearly invalidate this assumption, and would go absolutely undetected by our
+  algorithm, especially since it gives information over just a week of time.
+\end{itemize}
+
+These, however, are fundamental problems of the data we have available. They
+affect the results of the algorithm, but do not depend on it so they cannot be
+reduced using a more clever algorithm. We feel that, with the information
+available and in particular without an external source
+of standard routes, considering this assumption as true is the best
+possible approach.
+
+However, there are some issues that have to do with the particular
+implementation of the algorithm:
+
+\begin{itemize}
+  \item There could be situations where clusters have the same amount of routes,
+  or the numbers differ very slightly. In this cases, the accuracy of the
+  results inevitably decreases.
+  \item There could be areas that are poorly covered by the sensors, or that
+  have few flights passing though it. A cluster with five routes would be
+  selected against one with two or three, but clearly it does not have a strong
+  argument in determining the standard route between two airports. There are two
+  possible approaches to this. The first is to recognize that the informations
+  available is not enough, and possibly leave some pairs of airports with
+  undetermined standard routes. The other, and the one that we follow, is to
+  just run the algorithm without taking into account how many routes are
+  actually involved in the selection of the standard route. A possible
+  refinement could be to add to the front-end visualization of the standard
+  routes an indication of ``accuracy'' of a particular route based on the amount
+  of data available in the detection.
+\end{itemize}
+
+\paragraph{Clustering algorithm}
+
+Routes, both standard and nonstandard, are determined separately for each
+ordered pair of airports. The pairs are ordered in the sense that routes from
+Schiphol to Charles de Gaulle and Charles de Gaulle to Schiphol are considered
+separately. The algorithm proceeds by grouping together flights for each pair of
+airports. Then, given a particular pair, the following sequence of steps is
+executed for each flight, one after the other.
+
+\begin{itemize}
+  \item Test the flight agains every already present cluster, where the test is
+  positive if and only if the distance between the current flight and every
+  flight in the cluster is below a certain threshold
+  \item If the test is positive for some cluster, add the flight to that cluster
+  \item Otherwise, create a new cluster with the current flight as the only
+  classified flight
+\end{itemize}
+
+As it can be seen, the clustering algorithm proceeds by maintaining a list of
+clusters that gets updated at each iteration, making it fundamentally
+sequential. It follows that, on a large scale implementation, every execution of
+the clustering algorithm will be carried on by a single worker. The
+parallelization of the process, however, is not damaged: a separate clustering
+algorithm must be executed for each pair of airports, so it sufficies to
+distribute the data among the nodes by pair of airports.
+
+Our approach of clustering together routes represented as sequences of sectors
+is inspired by the Leader Algorithm described in [reference], but differs from
+it in some ways. Sectors in [reference] are bounded airspace regions under the
+control of a single air traffic controller or small team, hence they cover a
+relatively large area. Flights with such subdivision usually go through 5--10
+sectors. Our sectors are smaller --- each flight has approximately 20--40 of
+them --- therefore they represent a much more fine-grained subdivision of the
+world map.
+
+In [reference], distance between two routes is computed as a simple editing
+distance between a lexical representation of routes. With smaller sectors,
+however, comes the need to take other factors into account. An extreme example
+of why this is needed is given by two routes that are almost parallel and at
+distance of slightly more than 10 km. In our subdivision in sectors of 10 km by
+10 km, these two routes would end up represented by a sequence with almost no
+common sectors, even though they are actually very similar. A simple editing
+distance function would assign the same value to this pair of routes, as well as
+another pair of routes much more distant and different from each other.
+
+To avoid errors in the clustering, we also consider the euclidean distance
+between differing sectors of two routes as a weight in the final result. This
+values, together with the editing distance, determine the distance between two
+routes. The threshold at the heart of the algorithm, as well as the other
+parameters, have been fine-tuned with experimentations and
+observation of the results.
+
+\subsubsection{CO2 consumption}
+
+% ??????
 
@@
 
+% also add results of the different phases in terms of file size. for example,
+% how much the data shrinks from raw avro to position messages only.
+
 \section{Experiments}
 % a description of the experiments and their results.
+
+%%% position grouping
+% LOG: positions-20160918 has 199030 clusters and 1219 good clusters
+% LOG: positions-20160919 has 170594 clusters and 1103 good clusters
+% LOG: positions-20160920 has 126370 clusters and 1241 good clusters
+% LOG: positions-20160921 has 127855 clusters and 1205 good clusters
+% LOG: positions-20160922 has 127982 clusters and 1308 good clusters
+% LOG: positions-20160923 has 130662 clusters and 1243 good clusters
+% LOG: positions-20160924 has 121624 clusters and 874 good clusters
+
+%%% flight -> airline. how to, statistics: how many without airline
+
+%%% nondiverging vs diverging flights
 
@@
 
+% 70 active sensors in the opensky network
+
+\section{Future work}
+
+Improve algorithms:
+
+* flight detection
+  * dataset relating callsigns to flights
+  * using velocity and or more advanced techniques to make up for the misisng points
+* route detection
+  * use external database
+* CO2 consumption
+  * use icao to get aircraft characteristics
+
 \end{document}

### 2016-10-23 paper/paper.tex
PAPER: Add content

@@
 
-% Format of the output?
-
 \subsubsection{Flights}
@@
 particular, a group of points is considered a valid flight if and only if it
-starts and ends with points at an altitude of 3000 km or below, and shows an
+starts and ends with points at an altitude of 3000 m or below, and shows an
 ascent as well as a descent. Ascents and descents are defined as sequences of
 points which altitudes rise, respectively decrease, steadily for 2000 meters.
+
+We acknowledge that the criteria according to which we group and filter discrete
+flights are pretty restrictive, and that we inevitably throw away a lot of data
+and possibly some flights in the process. However, given the unreliability of
+the data and the difficult of our task, we wanted to deal with sufficiently
+precise and detailed data, in order to make our life easier in the subsequent
+phases of the processing. In particular, the restriction on the end points of a
+flight being below 3000 m is the most important one for our purposes (as
+explained below), and also the one that seems to filter out the majority of
+potential flights.
 
@@
 
-% ??????
+A little research showed that an average aircraft procudes about 53 pounds of
+carbon dioxide per mile. We used this value to compute the difference in CO2
+production between the standard routes and the corresponding straight-line
+route.
+
+We also wanted to offer a visualization of the comsumption of fuel with respect
+to a straight-line route. The visualization should show the points of a given
+route on the map in different colors on a palette from greed to red, where hues
+towards red would indicate a point of high consumption in the route.
+To do this, we needed to attribute to each segment of the route a suitable value,
+from which we could determine its color. Rather than trying to compute a value
+of ``comsumption'' for each segment, we decided to determine it based on
+how different the route is with respect to the straight-line route. The main
+characteristic of a straight-line route is that the airplane always points to
+the destination airport. Therefore, we considered as value of each segment the
+angle between the direction that the aircraft is facing in that segment and the
+direction of the destination airport. Higher values lead to more red segments.
+
+% TODO insert picture of route coloring here
 
@@
 
+\paragraph{Flight detection}
+
+We carried out some experiments on the results of the various phases and
+algorithms, in order to asses their behaviour. Some statistic data collected in
+the flight detection phase show how our classification criteria turned out to be
+pretty restrictive, as expected.
+
+This result should be attributed to the restrictive criteria of the
+classification algorithm, as well as to the incomplete information provided by
+the dataset. It should be recalled that, even though the OpenSky network
+comprises at the moment of 70 active sensors, some spots in the western Europe
+area may still be pooly covered, or not covered at all. Even a small poorly
+covered are makes all information on the flights passing through it incomplete,
+thus unsuitable for our analysis.
+
+It should also be considered that what our algorithm considers as a cluster is
+just a group of position points that is sufficiently dense. A cluster does not
+necessarily corresponds to a flight with incomplete data (and most of the times
+it doesn't). Comparing the number of detected flights with the total number of
+clusters is not a good indication of the undetected flights or the amound of
+position points discarded by the algorithm.
+Having said this, the table still gives an idea of how good the dataset is. In
+an ideal dataset, every cluster corresponds to a flight.
+
+% TODO make a table
 %%% position grouping
-% LOG: positions-20160918 has 199030 clusters and 1219 good clusters
-% LOG: positions-20160919 has 170594 clusters and 1103 good clusters
-% LOG: positions-20160920 has 126370 clusters and 1241 good clusters
-% LOG: positions-20160921 has 127855 clusters and 1205 good clusters
-% LOG: positions-20160922 has 127982 clusters and 1308 good clusters
-% LOG: positions-20160923 has 130662 clusters and 1243 good clusters
-% LOG: positions-20160924 has 121624 clusters and 874 good clusters
+% LOG: positions-20160918 has 199030 clusters and 1219 flights
+% LOG: positions-20160919 has 170594 clusters and 1103 flights
+% LOG: positions-20160920 has 126370 clusters and 1241 flights
+% LOG: positions-20160921 has 127855 clusters and 1205 flights
+% LOG: positions-20160922 has 127982 clusters and 1308 flights
+% LOG: positions-20160923 has 130662 clusters and 1243 flights
+% LOG: positions-20160924 has 121624 clusters and 874 flights
 
-%%% flight -> airline. how to, statistics: how many without airline
+\paragraph{Airlines}
+
+In order to compute the amount of flights diverging from a standard route per
+airline, we needed to associate a an airline to each flight. This is not a
+completely trivial task, as airline information is not provided at all by the
+data. Every aircraft, hence every flight, has an ICAO number associated to it,
+from which it is possible to determine callsigns from the identification
+messages in the dataset. In the case of commercial flights, the callsign of an
+aircraft contains the ICAO three-letter code of the airline for which the
+vehicle is flying for at the moment. Since aircraft are bought by airlines and
+hardly transferred from an airline to another, it is safe to assume that only
+one callsign is sufficient to determine the airline of an aircraft.
+The mapping between three-letter ICAO codes of airlines and their name has been
+done with the help of an external dataset.
+
+Of 19860 distinct icaos for which identification messages were available, 15911
+had a unique three-letter airline code in their callsigns, and 15691 had a code
+actually corresponding to an airline in the dataset. The remaining 4169
+icaos without a proper airline code are likely to correspond to private airplanes.
+
+\paragraph{Diverging flights}
 
@@
 
-% 70 active sensors in the opensky network
+... The OpenSky network counts 70 active sensors at the time of development,
+yielding a pretty good coverage of the western Europe area. Some areas are
+covered almost perfectly, and for those we were able to detect flight accurately.
+
+  * What are the standard routes?
+
+The results from the complete execution of the algorithm are quite good.
+It should be considered that there are a lot of minor airports, and flight
+information for those airports is, as one could expect, incomplete.
+Of 1165 of the total combinations of airports, about 85 \% of them has their
+standard route determined by under 10 flights. These pairs usually correspond
+to minor airports, and the accuracy of the computed standard route has
+necessarily to be considered pretty low. For major airports, however, the
+available data rises significantly. As an example, the standard route between
+Schiphol and London Gatwick and converse has been determined by 39,
+respectively 42 routes.
+
+% TODO say something about the distance function
+
+  * Which flights and airlines diverge from them? In which airports?
+  * How much CO2 would be saved if aircrafts followed a straight-line route
+    instead of the standard one?
+
+* Is it possible to determine flights from raw adsb data?
+
+Observing our results, we conclude that it is definitely possible to
+identify flights with acceptable reliability using raw ADS-B data alone. There
+are however some issues with the data that make this effort less than
+trivial, and some advanced techniques as well as other data sources may be
+needed to determine them with more than average accuracy.
 
@@
 
-Improve algorithms:
+We find that the accuracy of many of our algorithms is difficult to assess,
+given the lack of reliable data to compare against. Future work on this project
+could start by retrieving and using such external data, when available. The most
+significant effort would be to asses the correctness of the route clustering
+algorithm with a precompiled dataset of standard routes.
 
-* flight detection
-  * dataset relating callsigns to flights
-  * using velocity and or more advanced techniques to make up for the misisng points
-* route detection
-  * use external database
-* CO2 consumption
-  * use icao to get aircraft characteristics
+The next step would be to improve the algorithms, by making them more clever and
+use more data from the dataset in a better way. The flight detection algorithm
+could be improved by using an external dataset relating callsigns, which are
+amost fully available from the data, to flights. At the time of writing we are
+not aware of such dataset being publicly available, but there could be in the
+future. Also, our flight detection algorithm does not take velocity messages
+into account. Velocity information is not vital in flights detection, but we do
+not exclude that the heading information of an aircraft could be combined with
+its altitude information to better recognize ascents and descents.
+The grouping criteria in our flight detection are quite restrictive,
+to be able to get sufficiently reliable outputs with a relatively simple
+algorithm. These criteria can surely be made less restrictive, using some more
+advanced techniques to make up for missing points.
+
+There is room for improvement in the route detection algorithm, for example
+by tuning the way routes are clustered together, and in particular by selecting
+a better distance function. Another improvement would also come naturally by
+increasing the number of flights available to the analysis.
+A different approach could be to discard the current algorithm and instead use
+an external dataset of standard routes. In this setting, standard routes need
+not be computed. Flights that diverge from them can simply be discovered by
+computing the distance from the standard route for the particular pair of
+airports, and checking the result against a threshold.
+
+Currently, the CO2 consumption for the flights is computed in a very simple way,
+by multiplying the distance traveled by the airplanes by an estimate of the
+quantity of CO2 that an average aircraft produces per mile. A better approach
+could use the ICAO number of each single aircraft to index an external dataset
+and discover its characteristics. These properties could be used to compute a
+more precise and realistic estimate of CO2 production.
+
+% TODO: How other work could be based on ours?
 

### 2016-10-23 paper/paper.tex
PAPER: Add content

@@
 \subsection{OpenSky Network}
+
 Most of aircraft are provided with ADB-S transmitter, which send continuosly information about the status and the position of the aircraft to guarantee better air traffic control.
@@
 For these reasons OpenSky network was recently born.
-\\
+
 OpenSky is an open source network that aggregates ADS-B messages collected by volunteers around the globe.
@@
     The ADS-B standard defines different kinds of messages which are sent at different frequencies:
+
     \begin{itemize}
@@
     \end{itemize}
+
     In addition to these messages, other emergencies, priority, capability, navigation accuracy category, and operational modes messages could be broadcasted.
 
-\subsection{Route detection}
+\subsection{Routes detection}
+
 A way to reduce the number of air routes is illustrated in \textit{J.
@@
 The first prototype of this model has been developed to improve the air traffic flow in the National Airspace System of the USA.
+
+    \subsubsection{Routes distance}
+
+    Routes are represented by an ordered list of the geographical sectors where the aircrafts transit.
+Prior to define the clustering algorithm, it is important to define how to measure the distance between two different routes.
+Since each sector is identified by a code, the string editing distance can be used to measure the distance between two different routes.
+The distance is therefore represented by the minimum number of operation -- insertion, deletion and replacement -- to transform Route 1 into Route 2 or vice-versa.
+Different weight can also be applied to different operations.
+The distance should not depend by the length of the routes and for this reason it is then divided by the sum of the lengths of the routes.
+The final result represent the percentage of difference between the routes.
+
+    \subsubsection{Routes clustering}
+
+    Given a distance measure for each couple of routes, it is necessary to define a clustering algorithm to group them.
+Many methods has been tried during the research and the \textit{Leader algorithm} showed the best results.
+It proceed as follows:
+    \begin{enumerate}
+      \item A random route is assigned as leader of a new cluster;
+      \item All the other routes which distance from the leader is less than a threshold are assigned to the cluster;
+      \item Repeat the steps with the remaining routes until all routes are assigned to a cluster.
+    \end{enumerate}
 
@@
 
-\subsection{Algorithms}
+\subsection{Algorithm overview}
 
-\subsubsection{Positions}
+Our processing algorithm is composed of three phases that are executed in
+sequence. The output of a phase is the input of the next. The process starts
+from a dataset of raw ADS-B messages encoded in AVRO format, and ends producing
+the data necessary to answer our research questions. The phases are the
+following:
+
+\begin{enumerate}
+  \item \textbf{Position extraction}: the dataset of AVRO-encoded ADS-B messages
+  is analyzed to extract only the position information abount the aircrafts. In
+  particular, this phase produces a table of timestamped position points with
+  latitude, longitue and altitude. Every position point is associated to a
+  particular aircraft, identified by a unique ICAO code.
+  \item \textbf{Flight detection}: the unstructured position points are analyzed
+  per aircraft (per ICAO), and grouped together to extract single discrete
+  flights. For each detected flight, departure and arrival airports are
+  computed.
+  \item \textbf{Route detection}: flights are grouped by pair of departure and
+  arrival airport, and sent to an aggregative clustering algorithm that is used
+  to identify standard routes and routes that diverge from them.
+\end{enumerate}
+
+These three phases are explained in more detail in the sections that follow.
+
+\subsection{Positions}
 
@@
 
-\subsubsection{Flights}
-
-\paragraph{Identification of discrete flights}
+\subsection{Flights}
 
@@
 
-\paragraph{Departure and arrival airports}
+\subsubsection{Departure and arrival airports}
 
@@
 only commercial airports with the help of an external list we found on the
-Internet. This filtering reduced the list from XXX to XXX entries.
+Internet. This filtering reduced the list from 6977 to 639 entries.
 The resulting number of entries is still a little bit high for our purposes, but
@@
 been to build a dataset by hand, which seemed extremely time consuming.
-% TODO ^^^^^^^^^^
 
@@
 
-\paragraph{Sectors}
+\subsubsection{Sectors}
 
@@
 
-\subsubsection{Routes}
+\subsection{Routes}
 
@@
 
-\paragraph{Clustering algorithm}
+\subsubsection{Clustering algorithm}
 
@@
 observation of the results.
+
+To aid the visualization, we consider ``diverging clusters'' instead of
+diverging flights. After the clustering process, all clusters different from the
+one selected as standard are diverging clusters. We then display the
+representatives of the diverging clusters as the diverging routes for each
+airport. The rationale is simply that routes in the same cluster are close
+enough to be abstracted by a single representative of the entire group.
 
@@
 
-  * What are the standard routes?
-
-The results from the complete execution of the algorithm are quite good.
+The results from the complete execution of the algorithm are mixed.
 It should be considered that there are a lot of minor airports, and flight
@@
 
-% TODO say something about the distance function
-
-  * Which flights and airlines diverge from them? In which airports?
-  * How much CO2 would be saved if aircrafts followed a straight-line route
-    instead of the standard one?
-
-* Is it possible to determine flights from raw adsb data?
+Tuning the distance function for the route clustering phase was more challanging
+than we expected. In many cases, the algorithm ended up considering as diverging
+routes that were actually very close to the standard one. We can conclude,
+therefore, that even though we think we were able to determine standard routes
+with acceptable accuracy and confidence for medium to big airports, diverging
+routes detection still has some issues. We do think that the approach of using
+sectors and editing distance can give good results in this context, but the
+distance function obviously needs much more experiments are fine-tuning to be
+able to give good results in all situations.
 
@@
 The next step would be to improve the algorithms, by making them more clever and
-use more data from the dataset in a better way. The flight detection algorithm
-could be improved by using an external dataset relating callsigns, which are
-amost fully available from the data, to flights. At the time of writing we are
-not aware of such dataset being publicly available, but there could be in the
-future. Also, our flight detection algorithm does not take velocity messages
-into account. Velocity information is not vital in flights detection, but we do
-not exclude that the heading information of an aircraft could be combined with
-its altitude information to better recognize ascents and descents.
-The grouping criteria in our flight detection are quite restrictive,
-to be able to get sufficiently reliable outputs with a relatively simple
-algorithm. These criteria can surely be made less restrictive, using some more
-advanced techniques to make up for missing points.
+use more data from the dataset in a better way. A trivial but effective
+improvement to all phases of the algorithm follows, clearly, from increasing
+the number of flights available to them.
+
+The flight detection algorithm could be improved by using an external dataset
+relating callsigns, which are amost fully available from the data, to flights.
+At the time of writing we are not aware of such dataset being publicly
+available, but there could be in the future. Also, our flight detection
+algorithm does not take velocity messages into account. Velocity information is
+not vital in flights detection, but we do not exclude that the heading
+information of an aircraft could be combined with its altitude information to
+better recognize ascents and descents. The grouping criteria in our flight
+detection are quite restrictive, to be able to get sufficiently reliable outputs
+with a relatively simple algorithm. These criteria can surely be made less
+restrictive, using some more advanced techniques to make up for missing points.
 
@@
 by tuning the way routes are clustered together, and in particular by selecting
-a better distance function. Another improvement would also come naturally by
-increasing the number of flights available to the analysis.
+a better distance function.
+A possible alternative way to compute the distance could be to compute the area
+delimited by two routes. This area, normalized according to the length of the
+routes, could be a good estimate of how ``different'' two routes are.
 A different approach could be to discard the current algorithm and instead use

### 2016-10-23 paper/paper.tex
PAPER: add content

--- 2016-10-23 paper/paper.tex
@@
 \maketitle
-
-\begin{abstract}
-Abstract goes here.  
-\end{abstract}
 
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[scale=0.28]{imgs/4b7fad-altitudes.png}
+  \caption{Altitude values of the aircraft with ICAO 4b7fad, during September
+    20th, 2016. It can be seen that the data points disappear when the aircraft
+    approaches the ground. Also, in the group of points of the left, the
+    information on the first part of the flight is entirely missing, and the
+    first point available has timestamp 06:46 am. This is almost
+    certainly due to the airplane traveling through a poorly covered area.}
+  \label{fig:alts}
+\end{figure}
+
 Having extracted all position information from the raw avro files, there is now
@@
 \subsubsection{Departure and arrival airports}
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[scale=0.28]{imgs/4b7fad-altitudes-grouped.png}
+  \caption{Altitude values of the aircraft with ICAO 4b7fad, during September
+    20th, 2016, as grouped by the flight detection algorithm.}
+  \label{fig:alts}
+\end{figure}
 
@@
 
+\begin{figure}[t]
+  \centering
+  \includegraphics[scale=0.4]{imgs/amsterdam-vatry-co2.png}
+  \caption{Visualization of CO2 consumption for the standard route from
+    Amsterdam to Vatry. It can be seen that the route is mostly green for the
+    first part. It then shifts to red in the last part, where it takes a detour
+    from the straight-line route.}
+  \label{fig:amsterdam-vatry}
+\end{figure}
+
 A little research showed that an average aircraft procudes about 53 pounds of
@@
 
-% TODO insert picture of route coloring here
-
 \subsection{Deployment on the hadoop cluster}
@@
 
-% also add results of the different phases in terms of file size. for example,
-% how much the data shrinks from raw avro to position messages only.
+When working of large-scale datasets, it is essential to understand the
+available data and try to extract only relevant information from it. The
+subdivision of our algorithm in different phases is not only a convenient
+logical partitioning of operations, but it is essential to transform the dataset
+into smaller datasets, tailored to our needs. It is useful to do this
+transformation gradually, step by step, so that single intermediate phases can
+be reiterated without affecting the previous ones.
+
+The whole algorithm has been implemented as a Scala + Spark project, that has
+been deployed to the Hadoop cluster. We developed every phase of the algorithm
+as a separate executable Scala class, allowing us to run each phase separately.
+
+The position extraction phase was run on the entire dataset of raw AVRO-encoded
+messages, and it was, and one could expect, the most time consuming. The entire
+execution took about 48 hours. As a result of this phase, we were able to shrink
+the data from about 85 GB per day to about 10 GB of position points per day.
+This of course made the next phases of the algorithm more tractable. Position
+data has then been processed by the flight detection algorithm, in about 24
+hours. This phase, as we already explained, throws away a lot of data. Also,
+flights are represented and saved to disk as sequences of sectors, where each
+flight it composed of 20--50 sectors. This gives a very succint representation,
+and we were able to store the output of this phase in just about 4 MB per day.
+The small size of the flights data allowed us to run the route clustering
+algorithm and all the additional JSON files generation for the visualization in
+under 30 minutes. With such small execution time, we were able to test the
+results of the algorithm several times, and improve its performance and results.
 
@@
 % a description of the experiments and their results.
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[scale=0.4]{imgs/gatwick-schiphol.png}
+  \caption{Results of standard (blue) and diverging (red) routes detection for
+    flights from London Gatwick to Amsterdam Schiphol.}
+  \label{fig:gatwick-schiphol}
+\end{figure}
 
@@
 % tasks?
+
+\begin{figure}[t]
+  \centering
+  \includegraphics[scale=0.35]{imgs/schiphol-linate.png}
+  \caption{Results of standard (blue) and diverging (red) routes detection for
+    flights from Amsterdam Schiphol to Milano Linate. It can be seen that the
+    diverging route on the right is very similar to the standard route, but the
+    clustering algorithm fails to classify it as such.}
+  \label{fig:schiphol-linate}
+\end{figure}
 

### 2016-10-23 paper/paper.tex
Added introduction

@@
 
+The number of flying airplane has increased much in the last decade.
+This led to many problems, mostly related to security and air traffic controls.
+It was clear that radar surveillance systems would not be sufficient to monitor flights, since airplanes are passive entity.
+With the introduction of new technologies, a new standard called ADS-B has been developed to actively monitor aircrafts and manage routings.
+
+\textit{Automatic Dependant Surveillance - Broadcast} (ADS-B) is the last generation standard for air traffic monitoring.
+The main idea of this technology is that each aircraft broadcast its position and status using internal sensors, without relying on external entities.
+On the other side, each airplane is able to receive the messages and detect the position of other vehicle autonomously, thus maintening a minimum distance and avoiding collision.
+
+ADS-B messages are sent without any encryption and can be received and decoded by everyone.
+Morover this equipment will be mandatory in 2017 in most countries.
+We obtained a dataset, provided by OpenSky network, containing all the ADS-B messages sent by airplanes and collected by volunteers.
+The dataset includes data about flights from the 18\textsuperscript{th} September 2016 to the 24\textsuperscript{th} September 2016 and collected by about 70 sensors spread across Europe.
+
+In this paper we explain our analysis of such dataset, starting from the decoding of the position. The main goals of this project are:
+\begin{itemize}
+  \item Detecting the standard routes between all the pairs of airports;
+  \item Detecting abnormal routes between pairs of airports;
+  \item Calculate the wastage of fuel travelling through the standard route instead of flying over a straight line.
+\end{itemize}
+
+We will show that, though most standard routes are really close to the straight line, there are some routes that are clearly away from it.
+Moreover, for some couples of airports, there are few routes that differ from the standard one.
+
 \section{Related Work}
@@
 Most of aircraft are provided with ADB-S transmitter, which send continuosly information about the status and the position of the aircraft to guarantee better air traffic control.
-The messages are not encripted end, in most region, this technology will become mandatory in 2020.
+The messages are not encripted and, in most region, this technology will become mandatory in 2020.
 For these reasons OpenSky network was recently born.

### 2016-10-23 paper/paper.tex
PAPER: add content

@@
 
+With this project, we aim to answer the following questions:
+
 \begin{itemize}
-  \item What are the standard routes?
+  \item What are the standard routes between pairs of airports?
   \item Which flights and airlines diverge from them? In which airports?
@@
 
-Additionally
+To answer these questions, we are going to use a dataset of raw ADS-B messages
+encoded in AVRO format. This dataset provides information from the OpenSky
+network for a full week of September 2016. 
+The quantity of available messages is very high, but their content is missing
+some important informations about the flights, such as airlines and airports. It
+follows that there are other questions that must be faced in the development of
+the project:
 
@@
   \item How to use raw ADS-B data to identify flights?
+  \item How to use raw ADS-B to discover additional properties of flights, such
+  as associated airline or CO2 consumption of the aircraft?
+  \item How to identify the departure and arrival airports of a flight?
+  \item How to identify standard routes from an unstructured group of flights?
+  \item Is ADS-B data alone enough? Do we need external datasets?
 \end{itemize}
@@
 Our processing algorithm is composed of three phases that are executed in
-sequence. The output of a phase is the input of the next. The process starts
+sequence; the output of a phase is the input of the next. The process starts
 from a dataset of raw ADS-B messages encoded in AVRO format, and ends producing
@@
 
-... The OpenSky network counts 70 active sensors at the time of development,
+The OpenSky network counts 70 active sensors at the time of development,
 yielding a pretty good coverage of the western Europe area. Some areas are
 covered almost perfectly, and for those we were able to detect flight accurately.
+Observing our results, we conclude that it is definitely possible to
+identify flights with acceptable reliability using raw ADS-B data alone. There
+are however some issues with the data that make this effort less than
+trivial, and some advanced techniques as well as other data sources may be
+needed to determine them with more than average accuracy.
 
-The results from the complete execution of the algorithm are mixed.
+The results from the execution of the route detection algorithm are mixed.
 It should be considered that there are a lot of minor airports, and flight
@@
 Tuning the distance function for the route clustering phase was more challanging
-than we expected. In many cases, the algorithm ended up considering as diverging
-routes that were actually very close to the standard one. We can conclude,
+than we expected. In some cases, the algorithm ended up considering as diverging
+routes that were relatively close to the standard one. We can conclude,
 therefore, that even though we think we were able to determine standard routes
-with acceptable accuracy and confidence for medium to big airports, diverging
-routes detection still has some issues. We do think that the approach of using
-sectors and editing distance can give good results in this context, but the
-distance function obviously needs much more experiments are fine-tuning to be
-able to give good results in all situations.
-
-Observing our results, we conclude that it is definitely possible to
-identify flights with acceptable reliability using raw ADS-B data alone. There
-are however some issues with the data that make this effort less than
-trivial, and some advanced techniques as well as other data sources may be
-needed to determine them with more than average accuracy.
+with acceptable accuracy and confidence for medium- to big-size airports,
+diverging routes detection still has some issues. We do think that the approach
+of using sectors and editing distance can give good results in this context, but
+the distance function obviously needs more experiments are fine-tuning to
+be able to give good results in all situations.
 

### 2016-10-23 paper/paper.tex
Added subsection technologies

@@
 % python, haskell for scripting, etc
+
+Analyzing the entire dataset require a big computational power, therefore use a cluster to perform analyze the data.
+However, before running on the cluster, we tried the algorithms locally on a small subset of data to test correctness end efficiency and to obtain statistics.
+For this reason, we used different technologies depending on where we ran the analysis.
+
+On the cluster, we chose to use \textit{Apache Spark} because it is extremely fast while providing an easy interface, resulting suitable for our application.
+In particular, the \textit{Scala} version was chosen instead of PySpark because of the strongly typing and the functional programming approach.
+In addition to this, the Java library \textit{libadsb} was used to decode the position messages.
+
+Locally, in addition to \textit{Scala} we used \textit{Haskell} and \textit{Python} as scripting languages.
+The first one was suitable for data manipulation and algorithms testing, while the second one offers a convenient way to plot data using the library \textit{Matplotlib}.
 

### 2016-10-23 paper/paper.tex
Added cluster/flights per day table
Added bibliography
PAPER: indent

@@
 \usepackage{graphicx}
+\usepackage[style=numeric,backend=bibtex]{biblatex}
+
+\addbibresource{bibliography.bib}
 
@@
 
-The number of flying airplane has increased much in the last decade.
-This led to many problems, mostly related to security and air traffic controls.
-It was clear that radar surveillance systems would not be sufficient to monitor flights, since airplanes are passive entity.
-With the introduction of new technologies, a new standard called ADS-B has been developed to actively monitor aircrafts and manage routings.
+The number of flying airplane has increased much in the last decade. This led to
+many problems, mostly related to security and air traffic controls. It was clear
+that radar surveillance systems would not be sufficient to monitor flights,
+since airplanes are passive entity. With the introduction of new technologies, a
+new standard called ADS-B has been developed to actively monitor aircrafts and
+manage routings.
 
-\textit{Automatic Dependant Surveillance - Broadcast} (ADS-B) is the last generation standard for air traffic monitoring.
-The main idea of this technology is that each aircraft broadcast its position and status using internal sensors, without relying on external entities.
-On the other side, each airplane is able to receive the messages and detect the position of other vehicle autonomously, thus maintening a minimum distance and avoiding collision.
+\textit{Automatic Dependant Surveillance - Broadcast} (ADS-B) is the last
+generation standard for air traffic monitoring. The main idea of this technology
+is that each aircraft broadcast its position and status using internal sensors,
+without relying on external entities. On the other side, each airplane is able
+to receive the messages and detect the position of other vehicle autonomously,
+thus maintening a minimum distance and avoiding collision.
 
-ADS-B messages are sent without any encryption and can be received and decoded by everyone.
-Morover this equipment will be mandatory in 2017 in most countries.
-We obtained a dataset, provided by OpenSky network, containing all the ADS-B messages sent by airplanes and collected by volunteers.
-The dataset includes data about flights from the 18\textsuperscript{th} September 2016 to the 24\textsuperscript{th} September 2016 and collected by about 70 sensors spread across Europe.
+ADS-B messages are sent without any encryption and can be received and decoded
+by everyone. Morover this equipment will be mandatory in 2017 in most countries.
+We obtained a dataset, provided by OpenSky network, containing all the ADS-B
+messages sent by airplanes and collected by volunteers. The dataset includes
+data about flights from the 18\textsuperscript{th} September 2016 to the
+24\textsuperscript{th} September 2016 and collected by about 70 sensors spread
+across Europe.
 
-In this paper we explain our analysis of such dataset, starting from the decoding of the position. The main goals of this project are:
+In this paper we explain our analysis of such dataset, starting from the
+decoding of the position. The main goals of this project are:
 \begin{itemize}
@@
   \item Detecting abnormal routes between pairs of airports;
-  \item Calculate the wastage of fuel travelling through the standard route instead of flying over a straight line.
+  \item Calculate the wastage of fuel travelling through the standard route
+  instead of flying over a straight line.
 \end{itemize}
 
-We will show that, though most standard routes are really close to the straight line, there are some routes that are clearly away from it.
-Moreover, for some couples of airports, there are few routes that differ from the standard one.
+We will show that, though most standard routes are really close to the straight
+line, there are some routes that are clearly away from it. Moreover, for some
+couples of airports, there are few routes that differ from the standard one.
 
@@
 
-Most of aircraft are provided with ADB-S transmitter, which send continuosly information about the status and the position of the aircraft to guarantee better air traffic control.
-The messages are not encripted and, in most region, this technology will become mandatory in 2020.
-For these reasons OpenSky network was recently born.
+Most of aircraft are provided with ADB-S transmitter, which send continuosly
+information about the status and the position of the aircraft to guarantee
+better air traffic control. The messages are not encripted and, in most region,
+this technology will become mandatory in 2020. For these reasons OpenSky network
+was recently born.
 
-OpenSky is an open source network that aggregates ADS-B messages collected by volunteers around the globe.
-The dataset is accessible to everyone for non-profit purposes and the data is provided in the raw format, in order to perform research and statistics about air traffic.
-\\
-However there are some limitation to the data.
-Only part of Europe has a good coverage, which means there are many areas where is not yet possible to detect airplanes informations.
-Furthermore messages are collected by volunteers and there is not guarantee that antennas are working properly 24/7.
+OpenSky is an open source network that aggregates ADS-B messages collected by
+volunteers around the globe. The dataset is accessible to everyone for
+non-profit purposes and the data is provided in the raw format, in order to
+perform research and statistics about air traffic. \\
+However there are some limitation to the data. Only part of Europe has a good
+coverage, which means there are many areas where is not yet possible to detect
+airplanes informations. Furthermore messages are collected by volunteers and
+there is not guarantee that antennas are working properly 24/7.
 
@@
 
-    The ADS-B standard defines different kinds of messages which are sent at different frequencies:
+The ADS-B standard defines different kinds of messages which are sent at
+different frequencies:
 
 \begin{itemize}
-        \item \textit{Aircraft identification} messages contain the callsign of the airplane;
-        \item \textit{Airborne position} messages contain altitude, latititude and longitude of the airplane and are sent twice per second;
-        \item \textit{Airborne velocity} messages contain climb rate and velocity of the airplane.
+  \item \textit{Aircraft identification} messages contain the callsign of the
+  airplane;
+  \item \textit{Airborne position} messages contain altitude, latititude and
+  longitude of the airplane and are sent twice per second;
+  \item \textit{Airborne velocity} messages contain climb rate and velocity of
+  the airplane.
 \end{itemize}
 
-    In addition to these messages, other emergencies, priority, capability, navigation accuracy category, and operational modes messages could be broadcasted.
+In addition to these messages, other emergencies, priority, capability,
+navigation accuracy category, and operational modes messages could be
+broadcasted.
 
@@
 
-A way to reduce the number of air routes is illustrated in \textit{J.
-DeArmon et al, Air Route Clustering for a Queuing Network Model of the National Airspace System}.
-The research showed that similar routes can be grouped to reduce the number of path between a pair of airports thus decrease air traffic congestion.
-The same approach could be used to detect standard routes where aircraft fly, given the departure and arrival airports.
-The first prototype of this model has been developed to improve the air traffic flow in the National Airspace System of the USA.
+A way to reduce the number of air routes is illustrated in \textit{J. DeArmon et
+  al, Air Route Clustering for a Queuing Network Model of the National Airspace
+  System}. The research showed that similar routes can be grouped to reduce the
+number of path between a pair of airports thus decrease air traffic congestion.
+The same approach could be used to detect standard routes where aircraft fly,
+given the departure and arrival airports. The first prototype of this model has
+been developed to improve the air traffic flow in the National Airspace System
+of the USA.
 
@@
 
-    Routes are represented by an ordered list of the geographical sectors where the aircrafts transit.
-Prior to define the clustering algorithm, it is important to define how to measure the distance between two different routes.
-Since each sector is identified by a code, the string editing distance can be used to measure the distance between two different routes.
-The distance is therefore represented by the minimum number of operation -- insertion, deletion and replacement -- to transform Route 1 into Route 2 or vice-versa.
-Different weight can also be applied to different operations.
-The distance should not depend by the length of the routes and for this reason it is then divided by the sum of the lengths of the routes.
-The final result represent the percentage of difference between the routes.
+Routes are represented by an ordered list of the geographical sectors where the
+aircrafts transit. Prior to define the clustering algorithm, it is important to
+define how to measure the distance between two different routes. Since each
+sector is identified by a code, the string editing distance can be used to
+measure the distance between two different routes. The distance is therefore
+represented by the minimum number of operation -- insertion, deletion and
+replacement -- to transform Route 1 into Route 2 or vice-versa. Different weight
+can also be applied to different operations. The distance should not depend by
+the length of the routes and for this reason it is then divided by the sum of
+the lengths of the routes. The final result represent the percentage of
+difference between the routes.
 
@@
 
-    Given a distance measure for each couple of routes, it is necessary to define a clustering algorithm to group them.
-Many methods has been tried during the research and the \textit{Leader algorithm} showed the best results.
-It proceed as follows:
+Given a distance measure for each couple of routes, it is necessary to define a
+clustering algorithm to group them. Many methods has been tried during the
+research and the \textit{Leader algorithm} showed the best results. It proceed
+as follows:
+
 \begin{enumerate}
   \item A random route is assigned as leader of a new cluster;
-      \item All the other routes which distance from the leader is less than a threshold are assigned to the cluster;
-      \item Repeat the steps with the remaining routes until all routes are assigned to a cluster.
+  \item All the other routes which distance from the leader is less than a
+  threshold are assigned to the cluster;
+  \item Repeat the steps with the remaining routes until all routes are assigned
+  to a cluster.
 \end{enumerate}
@@
 
-Analyzing the entire dataset require a big computational power, therefore use a cluster to perform analyze the data.
-However, before running on the cluster, we tried the algorithms locally on a small subset of data to test correctness end efficiency and to obtain statistics.
-For this reason, we used different technologies depending on where we ran the analysis.
+Analyzing the entire dataset require a big computational power, therefore use a
+cluster to perform analyze the data. However, before running on the cluster, we
+tried the algorithms locally on a small subset of data to test correctness end
+efficiency and to obtain statistics. For this reason, we used different
+technologies depending on where we ran the analysis.
 
-On the cluster, we chose to use \textit{Apache Spark} because it is extremely fast while providing an easy interface, resulting suitable for our application.
-In particular, the \textit{Scala} version was chosen instead of PySpark because of the strongly typing and the functional programming approach.
-In addition to this, the Java library \textit{libadsb} was used to decode the position messages.
+On the cluster, we chose to use \textit{Apache Spark} because it is extremely
+fast while providing an easy interface, resulting suitable for our application.
+In particular, the \textit{Scala} version was chosen instead of PySpark because
+of the strongly typing and the functional programming approach. In addition to
+this, the Java library \textit{libadsb} was used to decode the position
+messages.
 
-Locally, in addition to \textit{Scala} we used \textit{Haskell} and \textit{Python} as scripting languages.
-The first one was suitable for data manipulation and algorithms testing, while the second one offers a convenient way to plot data using the library \textit{Matplotlib}.
+Locally, in addition to \textit{Scala} we used \textit{Haskell} and
+\textit{Python} as scripting languages. The first one was suitable for data
+manipulation and algorithms testing, while the second one offers a convenient
+way to plot data using the library \textit{Matplotlib}.
 
@@
 the flight detection phase show how our classification criteria turned out to be
-pretty restrictive, as expected.
+pretty restrictive, as expected. The collected data is showed in Table~\ref{table}.
 
@@
 %%% position grouping
-% LOG: positions-20160918 has 199030 clusters and 1219 flights
-% LOG: positions-20160919 has 170594 clusters and 1103 flights
-% LOG: positions-20160920 has 126370 clusters and 1241 flights
-% LOG: positions-20160921 has 127855 clusters and 1205 flights
-% LOG: positions-20160922 has 127982 clusters and 1308 flights
-% LOG: positions-20160923 has 130662 clusters and 1243 flights
-% LOG: positions-20160924 has 121624 clusters and 874 flights
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c|c|}
+  \hline
+  \textbf{Day in dataset} & \textbf{Clusters} & \textbf{Detected flights} \\
+  \hline
+  18/09/2016 & 199030 & 1219 \\
+  \hline
+  19/09/2016 & 170594 & 1103 \\
+  \hline
+  20/09/2016 & 126370 & 1241 \\
+  \hline
+  21/09/2016 & 127855 & 1205 \\
+  \hline
+  22/09/2016 & 127982 & 1308 \\
+  \hline
+  23/09/2016 & 130662 & 1243 \\
+  \hline
+  24/09/2016 & 121624 & 874 \\
+  \hline
+  \end{tabular}
+  \caption{Number of clusters and flights per day}
+  \label{table}
+\end{table}
 
@@
 
+\printbibliography
+
 \end{document}

### 2016-10-23 paper/paper.tex
Updated authors
PAPER: stuff

@@
 \title{Title goes here}
-\author{Filippo Sestini \\ Vrije Universiteit \\ email@host.com \and
-  Davide Dal Bianco \\ Vrije Universiteit \\ email@host.com}
+\author{Filippo Sestini \\ Vrije Universiteit Amsterdam \\ f.sestini@student.vu.nl \and
+  Davide Dal Bianco \\ Vrije Universiteit Amsterdam \\ d.dalbianco@student.vu.nl}
 
@@
 \section{Introduction}
-% what is this project about and why is it interesting?
 
@@
 Our approach of clustering together routes represented as sequences of sectors
-is inspired by the Leader Algorithm described in [reference], but differs from
-it in some ways. Sectors in [reference] are bounded airspace regions under the
-control of a single air traffic controller or small team, hence they cover a
-relatively large area. Flights with such subdivision usually go through 5--10
-sectors. Our sectors are smaller --- each flight has approximately 20--40 of
-them --- therefore they represent a much more fine-grained subdivision of the
-world map.
+is inspired by the Leader Algorithm described in \cite{routeclustering}, but
+differs from it in some ways. Sectors in \cite{routeclustering} are bounded
+airspace regions under the control of a single air traffic controller or small
+team, hence they cover a relatively large area. Flights with such subdivision
+usually go through 5--10 sectors. Our sectors are smaller --- each flight has
+approximately 20--40 of them --- therefore they represent a much more
+fine-grained subdivision of the world map.
 
-In [reference], distance between two routes is computed as a simple editing
-distance between a lexical representation of routes. With smaller sectors,
-however, comes the need to take other factors into account. An extreme example
-of why this is needed is given by two routes that are almost parallel and at
-distance of slightly more than 10 km. In our subdivision in sectors of 10 km by
-10 km, these two routes would end up represented by a sequence with almost no
-common sectors, even though they are actually very similar. A simple editing
+In \cite{routeclustering}, distance between two routes is computed as a simple
+editing distance between a lexical representation of routes. With smaller
+sectors, however, comes the need to take other factors into account. An extreme
+example of why this is needed is given by two routes that are almost parallel
+and at distance of slightly more than 10 km. In our subdivision in sectors of 10
+km by 10 km, these two routes would end up represented by a sequence with almost
+no common sectors, even though they are actually very similar. A simple editing
 distance function would assign the same value to this pair of routes, as well as

### 2016-10-23 paper/paper.tex
Added citations

@@
 this technology will become mandatory in 2020. For these reasons OpenSky network
-was recently born.
+was recently born\cite{opensky}.
 
@@
 
-A way to reduce the number of air routes is illustrated in \textit{J. DeArmon et
-  al, Air Route Clustering for a Queuing Network Model of the National Airspace
-  System}. The research showed that similar routes can be grouped to reduce the
-number of path between a pair of airports thus decrease air traffic congestion.
-The same approach could be used to detect standard routes where aircraft fly,
-given the departure and arrival airports. The first prototype of this model has
-been developed to improve the air traffic flow in the National Airspace System
-of the USA.
+A way to reduce the number of air routes is illustrated bt J. DeArmon.
+The research showed that similar routes can be grouped to reduce the
+number of path between a pair of airports thus decrease air traffic 
+congestion\cite{routeclustering}. The same approach could be used to detect
+standard routes where aircraft fly, given the departure and arrival airports.
+The first prototype of this model has been developed to improve the air traffic
+flow in the National Airspace System of the USA.
 

### 2016-10-23 paper/paper.tex
asd

@@
 
-The number of flying airplane has increased much in the last decade. This led to
-many problems, mostly related to security and air traffic controls. It was clear
+The number of flying airplanes has increased much in the last decade. This led to
+many problems, mostly related to security and air traffic control. It was clear
 that radar surveillance systems would not be sufficient to monitor flights,
-since airplanes are passive entity. With the introduction of new technologies, a
+since airplanes are passive entities. With the introduction of new technologies, a
 new standard called ADS-B has been developed to actively monitor aircrafts and
@@
 generation standard for air traffic monitoring. The main idea of this technology
-is that each aircraft broadcast its position and status using internal sensors,
-without relying on external entities. On the other side, each airplane is able
-to receive the messages and detect the position of other vehicle autonomously,
-thus maintening a minimum distance and avoiding collision.
+is that each aircraft broadcasts its position and status using internal sensors,
+without relying on external entities. On the other hand, each airplane is able
+to receive the messages and detect the position of other vehicles autonomously,
+thus allowing it to maintain a minimum distance and avoid collision.
 
 ADS-B messages are sent without any encryption and can be received and decoded
-by everyone. Morover this equipment will be mandatory in 2017 in most countries.
-We obtained a dataset, provided by OpenSky network, containing all the ADS-B
-messages sent by airplanes and collected by volunteers. The dataset includes
-data about flights from the 18\textsuperscript{th} September 2016 to the
-24\textsuperscript{th} September 2016 and collected by about 70 sensors spread
-across Europe.
-
-In this paper we explain our analysis of such dataset, starting from the
-decoding of the position. The main goals of this project are:
-\begin{itemize}
-  \item Detecting the standard routes between all the pairs of airports;
-  \item Detecting abnormal routes between pairs of airports;
-  \item Calculate the wastage of fuel travelling through the standard route
-  instead of flying over a straight line.
-\end{itemize}
-
-We will show that, though most standard routes are really close to the straight
-line, there are some routes that are clearly away from it. Moreover, for some
-couples of airports, there are few routes that differ from the standard one.
+by anyone. We were able to obtain a dataset, provided by OpenSky network,
+containing all the ADS-B messages sent by airplanes and collected by volunteers
+during a time period spanning a week. The dataset includes data from September
+18\textsuperscript{th} 2016 to September 24\textsuperscript{th} 2016, as
+collected by about 70 sensors spread across western Europe. It is the purpose of
+this paper to explain our approach in analyzing and extracting data from such
+dataset. It is particularly intresting to see to which point data like the one
+offered by OpenSky is suitable to get useful information about flights. In this
+project, we concentrate on analyzing the air traffic and its patterns, and in
+particular why try to discover standard routes between pairs of
+airports, as well as flights that diverge from them.
 
 \section{Related Work}
-% meaningfully summarize related literature to explain what this project builds
-% on, in particular, discuss the two research papers chosen with this project
-% (and discussed in your presentation)
 
@@
 \section{Research Questions}
-% which questions is this project trying to answer, and/or hypotheses to
-% investigate? These should cover both the project topic, as well as the
-% technological side (aptness of the tools for the job, scalability of the
-% solution).
 
@@
 
-Analyzing the entire dataset require a big computational power, therefore use a
-cluster to perform analyze the data. However, before running on the cluster, we
-tried the algorithms locally on a small subset of data to test correctness end
-efficiency and to obtain statistics. For this reason, we used different
-technologies depending on where we ran the analysis.
+Analyzing the entire dataset requires a lot of computational power, hence the
+need to use a cluster to perform analysis on the data. However, before executing
+on the cluster, it is essential to be able to try the algorithms locally on a
+small subset of data, in order to test their correctness end efficiency, and to
+obtain useful statistics. For this reason, we used different technologies
+depending on where we ran the analysis.
 
 On the cluster, we chose to use \textit{Apache Spark} because it is extremely
-fast while providing an easy interface, resulting suitable for our application.
-In particular, the \textit{Scala} version was chosen instead of PySpark because
-of the strongly typing and the functional programming approach. In addition to
-this, the Java library \textit{libadsb} was used to decode the position
+fast while providing an easy and convenient development interface. In
+particular, the \textit{Scala} version was chosen instead of PySpark because of
+the strongly typing and the functional programming approach. In addition to
+this, the Java library \textit{libadsb} was used to decode the raw ADS-B
 messages.
 
-Locally, in addition to \textit{Scala} we used \textit{Haskell} and
-\textit{Python} as scripting languages. The first one was suitable for data
-manipulation and algorithms testing, while the second one offers a convenient
-way to plot data using the library \textit{Matplotlib}.
+On our local datasets, in addition to \textit{Scala} we used \textit{Haskell}
+and \textit{Python} as scripting languages for fast prototyping. The first one
+was suitable for data manipulation and algorithms testing, while the second one
+offers a convenient way to plot data using the library \textit{Matplotlib}.
 
@@
 
-\paragraph{Flight detection}
+\subsection{Flight detection}
 
@@
 
-\paragraph{Airlines}
+\subsection{Airlines}
 
@@
 
-\paragraph{Diverging flights}
+\subsection{Diverging flights}
 
@@
   \caption{Results of standard (blue) and diverging (red) routes detection for
-    flights from Amsterdam Schiphol to Milano Linate. It can be seen that the
-    diverging route on the right is very similar to the standard route, but the
-    clustering algorithm fails to classify it as such.}
+    flights from Amsterdam Schiphol to Milano Linate, during our previous
+    experiments. It can be seen that the diverging route on the right is very
+    similar to the standard route, but the clustering algorithm fails to
+    classify it as such. This made us increase the threshold in the
+    classification of routes in clusters.}
   \label{fig:schiphol-linate}
@@
 
-The results from the execution of the route detection algorithm are mixed.
+The results from the execution of the route detection algorithm are quite good.
 It should be considered that there are a lot of minor airports, and flight
-information for those airports is, as one could expect, incomplete.
-Of 1165 of the total combinations of airports, about 85 \% of them has their
-standard route determined by under 10 flights. These pairs usually correspond
-to minor airports, and the accuracy of the computed standard route has
-necessarily to be considered pretty low. For major airports, however, the
-available data rises significantly. As an example, the standard route between
-Schiphol and London Gatwick and converse has been determined by 39,
-respectively 42 routes.
+information for those airports is, as one could expect, incomplete. Of 1165 of
+the total combinations of airports, about 85 \% of them has their standard route
+determined by under 10 flights. These pairs usually correspond to minor
+airports, and the accuracy of the computed standard route has necessarily to be
+considered pretty low. For major airports, however, the available data rises
+significantly. As an example, the standard route between Schiphol and London
+Gatwick and converse has been determined by 39, respectively 42, routes. This
+gives more confidence on the accuracy of the result.
 

### 2016-10-23 paper/paper.tex
PAPER: stuff

@@
 
-\title{Title goes here}
+\title{Title}
 \author{Filippo Sestini \\ Vrije Universiteit Amsterdam \\ f.sestini@student.vu.nl \and
@@
 
-Most of aircraft are provided with ADB-S transmitter, which send continuosly
-information about the status and the position of the aircraft to guarantee
-better air traffic control. The messages are not encripted and, in most region,
+Most aircrafts have equipped with an ADB-S transmitter, with which they send
+information about their status and the position continuously, to guarantee
+better air traffic control. These messages are not encripted and, in most regions,
 this technology will become mandatory in 2020. For these reasons OpenSky network
-was recently born\cite{opensky}.
+has been created \cite{opensky}.
 
@@
 volunteers around the globe. The dataset is accessible to everyone for
-non-profit purposes and the data is provided in the raw format, in order to
+non-profit purposes and the data is provided in raw format, in order to
 perform research and statistics about air traffic. \\
-However there are some limitation to the data. Only part of Europe has a good
-coverage, which means there are many areas where is not yet possible to detect
-airplanes informations. Furthermore messages are collected by volunteers and
-there is not guarantee that antennas are working properly 24/7.
+There are some limitation to the data that is collected by the network. Western
+Europe is well, but only partially covered. In areas where coverage is spotty or
+absent, it is not possible to detect complete airplanes informations.
+Additionally, messages are collected by volunteers and there are no guarantees
+about the reliability of the collected information.
 
@@
 The ADS-B standard defines different kinds of messages which are sent at
-different frequencies:
+different frequencies. Among those types, four are particularly useful to
+extract information about airplanes and flights:
 
@@
 
-In addition to these messages, other emergencies, priority, capability,
-navigation accuracy category, and operational modes messages could be
-broadcasted.
+\subsection{Route clustering}
 
-\subsection{Routes detection}
-
-A way to reduce the number of air routes is illustrated bt J. DeArmon.
-The research showed that similar routes can be grouped to reduce the
-number of path between a pair of airports thus decrease air traffic 
-congestion\cite{routeclustering}. The same approach could be used to detect
-standard routes where aircraft fly, given the departure and arrival airports.
-The first prototype of this model has been developed to improve the air traffic
-flow in the National Airspace System of the USA.
+The research led by J. DeArmon showed that similar routes can be grouped
+together based on their similarity, and the result of this grouping can be used
+to reduce the number of paths between a pair of airports, thus decreasing air
+traffic congestion\cite{routeclustering}. The first prototype of this model has
+been developed to improve the air traffic flow in the National Airspace System
+of the USA.
 
@@
 Routes are represented by an ordered list of the geographical sectors where the
-aircrafts transit. Prior to define the clustering algorithm, it is important to
+aircrafts transit. Prior to devising the clustering algorithm, it is important to
 define how to measure the distance between two different routes. Since each
@@
 measure the distance between two different routes. The distance is therefore
-represented by the minimum number of operation -- insertion, deletion and
-replacement -- to transform Route 1 into Route 2 or vice-versa. Different weight
+represented by the minimum number of operations -- insertion, deletion and
+replacement -- to transform Route 1 into Route 2 or vice-versa. Different weights
 can also be applied to different operations. The distance should not depend by
-the length of the routes and for this reason it is then divided by the sum of
-the lengths of the routes. The final result represent the percentage of
-difference between the routes.
+the length of the routes and for this reason it is divided by the sum of
+the lengths of the routes. The final result represents the percentage of
+difference between the routes\cite{routeclustering}.
 
@@
 
-Given a distance measure for each couple of routes, it is necessary to define a
-clustering algorithm to group them. Many methods has been tried during the
-research and the \textit{Leader algorithm} showed the best results. It proceed
-as follows:
+Given a suitable distance measure for each couple of routes, it is necessary to
+define a clustering algorithm to group them. Many methods have been tried during
+the research described in \cite{routeclustering}, and the \textit{Leader
+  algorithm} showed the best results. The algorithm proceeds by processing the
+entire sequence of routes to be clustered, and it starts by assigning the first
+of these routes to the first cluster. In then examines each route in turn: a
+route that is sufficiently close to the leader of a cluster is assigned to it.
+If no leader is found to be close enough to the current route, it is assigned as
+the leader of a new cluster.
 
-\begin{enumerate}
-  \item A random route is assigned as leader of a new cluster;
-  \item All the other routes which distance from the leader is less than a
-  threshold are assigned to the cluster;
-  \item Repeat the steps with the remaining routes until all routes are assigned
-  to a cluster.
-\end{enumerate}
+As it is explained in detail below, our route detection algorithm comprises a
+part of route clustering that is inspired in many ways by the above algorithm,
+as described in \cite{routeclustering}.
 
@@
 
-% TODO: How other work could be based on ours?
-
 \printbibliography

### 2016-10-23 paper/paper.tex
Title

@@
 
-\title{Title}
+\title{Flight clustering for route detection and CO{\Large{\textsubscript{2}}} emissions}
 \author{Filippo Sestini \\ Vrije Universiteit Amsterdam \\ f.sestini@student.vu.nl \and
@@
 project, we concentrate on analyzing the air traffic and its patterns, and in
-particular why try to discover standard routes between pairs of
+particular we try to discover standard routes between pairs of
 airports, as well as flights that diverge from them.

### 2016-10-23 visualization/public/airlines.js
Updated routes

@@ -1,1 +1,123 @@
-var airlines = [{"name":"Aerodienst GmbH","nDiverging":1},{"name":"Airbus Industrie","nDiverging":3},{"name":"Air Berlin","nDiverging":16},{"name":"Austrian Airlines","nDiverging":1},{"name":"Alitalia","nDiverging":23},{"name":"Air Transat","nDiverging":1},{"name":"British Airways","nDiverging":2},{"name":"Croatia Airlines","nDiverging":2},{"name":"Czech Airlines","nDiverging":6},{"name":"Ethiopian Airlines","nDiverging":1},{"name":"European Air Transport","nDiverging":22},{"name":"Eurowings","nDiverging":1},{"name":"easyJet","nDiverging":15},{"name":"Federal Express","nDiverging":2},{"name":"Germanwings","nDiverging":20},{"name":"KLM Royal Dutch Airlines","nDiverging":13},{"name":"Lufthansa","nDiverging":42},{"name":"Lufthansa Cargo","nDiverging":1},{"name":"Luxair","nDiverging":3},{"name":"MNG Airlines","nDiverging":1},{"name":"Neos","nDiverging":1},{"name":"Qatar Airways","nDiverging":1},{"name":"Ryanair","nDiverging":24},{"name":"Swiss International Air Lines","nDiverging":14},{"name":"Swissair","nDiverging":14},{"name":"Stuttgarter Flugdienst","nDiverging":1},{"name":"Thomsonfly","nDiverging":1},{"name":"Tiphook PLC","nDiverging":2},{"name":"United Parcel Service","nDiverging":1},{"name":"Vistajet","nDiverging":1},{"name":"Vueling Airlines","nDiverging":9},{"name":"Wizz Air","nDiverging":6}]
+var airlines = 
+[
+    {
+        "diverging": 3,
+        "name": "Luxair",
+        "nonDiverging": 24
+    },
+    {
+        "diverging": 1,
+        "name": "Neos",
+        "nonDiverging": 5
+    },
+    {
+        "diverging": 1,
+        "name": "Qatar Airways",
+        "nonDiverging": 3
+    },
+    {
+        "diverging": 20,
+        "name": "Ryanair",
+        "nonDiverging": 124
+    },
+    {
+        "diverging": 38,
+        "name": "Swiss International Air Lines",
+        "nonDiverging": 794
+    },
+    {
+        "diverging": 3,
+        "name": "Airbus Industrie",
+        "nonDiverging": 2
+    },
+    {
+        "diverging": 7,
[…]

### 2016-10-23 paper/paper.tex
PAPER: add content

@@
 \end{itemize}
+
+Every ADS-B transmitter is assigned a unique 24-bit ICAO number. Since the
+transmitters installed on the airplanes are hardly changed, we can safely
+consider an aircraft and its ADS-B transmitter as the same entity. Every raw
+ADS-B message records the ICAO number of the transmitter, hence the aircraft,
+that sent it. This is basically the way we are able to associate information
+from the ADS-B messages to individual airplanes.
 
@@
   \end{tabular}
-  \caption{Number of clusters and flights per day}
+  \caption{Amount of clusters and flights per day, as determined by the flight
+    detection algorithm}
   \label{table}
@@
 messages in the dataset. In the case of commercial flights, the callsign of an
-aircraft contains the ICAO three-letter code of the airline for which the
+aircraft contains the three-letter code of the airline for which the
 vehicle is flying for at the moment. Since aircraft are bought by airlines and
@@
 one callsign is sufficient to determine the airline of an aircraft.
-The mapping between three-letter ICAO codes of airlines and their name has been
+The mapping between three-letter codes of airlines and their name has been
 done with the help of an external dataset.
 
-Of 19860 distinct icaos for which identification messages were available, 15911
+Of 19860 distinct ICAOs for which identification messages were available, 15911
 had a unique three-letter airline code in their callsigns, and 15691 had a code
 actually corresponding to an airline in the dataset. The remaining 4169
-icaos without a proper airline code are likely to correspond to private airplanes.
+ICAOs without a proper airline code are likely to correspond to private airplanes.
 
@@
 %%% nondiverging vs diverging flights
+
+We were able to associate an airline to almost all of the flights the resulted
+from the route clustering phase. From this information, we then determined how
+many flights, per airline, follow and do not follow the standard route for their
+particular pair of airports, as identified by our standard route detection.
+We observed that the majority of flights do follow tha standard route for their
+particular pair of departure and arrival airports. Table~\ref{airline-table}
+shows an excerpt of the data that we were able to collect.
+
+\begin{table}
+  \centering
+  \begin{tabular}{|c|c|c|}
+  \hline
+  \textbf{Airline} & \textbf{Normal} & \textbf{Diverging} \\
+  \hline
+  KLM Royal Dutch Airlines & 203 & 14 \\
+  Ryanair & 124 & 20 \\
+  Swiss International Air Lines & 794 & 38 \\
+  Alitalia & 412 & 39 \\
+  British Airways & 12 & 2 \\
+  Lufthansa & 1329 & 38 \\
+  easyJet & 108 & 15 \\
+  Germanwings & 191 & 18 \\
+  \hline
+  \end{tabular}
+  \caption{Excerpt of data relating some major airlines to flights that comply
+    or diverge from a standard route}
+  \label{airline-table}
+\end{table}
+
+From these results we may conclude that, especially for big companies with many
+flights associated to them, the standard routes as detected by our algorithm are
+really ``standard'', in the sense that there is a significant difference between
+flights that follow them and flights that do not.
 

### 2016-10-23 paper/paper.tex
paper

@@
 
-Most aircrafts have equipped with an ADB-S transmitter, with which they send
-information about their status and the position continuously, to guarantee
+Most aircrafts are equipped with an ADB-S transmitter, with which they send
+information about their status and position continuously, to guarantee
 better air traffic control. These messages are not encripted and, in most regions,
@@
 OpenSky is an open source network that aggregates ADS-B messages collected by
-volunteers around the globe. The dataset is accessible to everyone for
+volunteers around the globe. The dataset is accessible by everyone for
 non-profit purposes and the data is provided in raw format, in order to
@@
 Europe is well, but only partially covered. In areas where coverage is spotty or
-absent, it is not possible to detect complete airplanes informations.
+absent, it is not possible to detect complete airplane information.
 Additionally, messages are collected by volunteers and there are no guarantees
@@
 replacement -- to transform Route 1 into Route 2 or vice-versa. Different weights
-can also be applied to different operations. The distance should not depend by
+can also be applied to different operations. The distance should not depend on
 the length of the routes and for this reason it is divided by the sum of
@@
 encoded in AVRO format. This dataset provides information from the OpenSky
-network for a full week of September 2016. 
-The quantity of available messages is very high, but their content is missing
-some important informations about the flights, such as airlines and airports. It
-follows that there are other questions that must be faced in the development of
-the project:
+network for a full week of September 2016. The quantity of available messages is
+very high, but their content is missing some important informations about the
+flights, such as airlines and airports. It follows that there are other
+questions that must be answered in the development of the project:
 
@@
   \item How to use raw ADS-B data to identify flights?
-  \item How to use raw ADS-B to discover additional properties of flights, such
+  \item How to use raw ADS-B data to discover additional properties of flights, such
   as associated airline or CO2 consumption of the aircraft?
@@
   \item How to identify standard routes from an unstructured group of flights?
-  \item Is ADS-B data alone enough? Do we need external datasets?
+  \item Is ADS-B data alone enough to answer the above questions? Do we need
+  external datasets?
 \end{itemize}
@@
 Analyzing the entire dataset requires a lot of computational power, hence the
-need to use a cluster to perform analysis on the data. However, before executing
-on the cluster, it is essential to be able to try the algorithms locally on a
-small subset of data, in order to test their correctness end efficiency, and to
-obtain useful statistics. For this reason, we used different technologies
-depending on where we ran the analysis.
+need to use a cluster to perform the computations on the entire dataset.
+However, before executing on the cluster, it is essential to be able to try the
+algorithms locally on a small subset of the data, in order to test their correctness
+end efficiency, and to obtain useful statistics.
 
@@
 particular, the \textit{Scala} version was chosen instead of PySpark because of
-the strongly typing and the functional programming approach. In addition to
+the strong typing and the functional programming approach. In addition to
 this, the Java library \textit{libadsb} was used to decode the raw ADS-B
@@
   \item Filter out messages that are not position messages
-  \item Order the messages according to their timestamp
   \item Group the messages by ICAO
+  \item For each ICAO, order the messages according to their timestamp
   \item For each ICAO, go through the messages in order of timestamp and decode
@@
 From our observations, however, we concluded that these flights are not very
-frequent, so at the end of the day the final results did not suffer too much.
+frequent, so the final results did not suffer too much.
 
@@
 
-Having extracted all position information from the raw avro files, there is now
+Having extracted all position information from the raw AVRO files, there is now
 the problem of identifying discrete flights from it.
@@
 simple clustering algorithm: all points are analyzed in order of timestamp, and
-two subsequent points are considered belonging in the same cluster if and only
+two subsequent points are considered belonging to the same cluster if and only
 if they are no more than 20 minutes apart, their distance is no more than 20
@@
 and possibly some flights in the process. However, given the unreliability of
-the data and the difficult of our task, we wanted to deal with sufficiently
-precise and detailed data, in order to make our life easier in the subsequent
-phases of the processing. In particular, the restriction on the end points of a
-flight being below 3000 m is the most important one for our purposes (as
-explained below), and also the one that seems to filter out the majority of
-potential flights.
+the data and the difficulty of our task following from it, we wanted to deal
+with sufficiently precise and detailed data, in order to make our life easier in
+the next phases of the processing. In particular, the restriction on the end
+points of a flight being below 3000 m is the most important one for our purposes
+(as explained below), and unfortunately also the one that seems to filter out
+the majority of potential flights.
 
@@
 representation of flights and routes as a sequence of sectors, where each sector
-represents portion of the world map. For this reason, the flights detection
+represents a portion of the world map. For this reason, the flights detection
 phase outputs flights as sequences of sectors rather than position points. This
@@
 reduce the amount of data needed to represent a single flight, but the routes
-detection algorithm would stil have transformed those representations to
+detection algorithm would still have transformed those representations to
 sequences of sectors. It follows that converting directly from position points
@@
 sectors by simply truncating the latitude and longitude part of a point to one
-decimal digit. Strings of values with the same truncated latitude and
+decimal digit. Contiguous sequences of values with the same truncated latitude and
 longitude represent the same sector, and are compressent into one. This process
@@
 The most important part of the process is the detection of routes between pairs
-of airports. Standard routes between a pair airports are likely to be used by
+of airports. Standard routes between a pair of airports are likely to be used by
 the majority of flights between those airports. Our algorithm relies on the
@@
   \item Select a route that constitues the representative of
-  that cluster as the one having the least distance between al the other
+  that cluster as the one having the least distance between all the other
   routes in the cluster
@@
   clearly invalidate this assumption, and would go absolutely undetected by our
-  algorithm, especially since it gives information over just a week of time.
+  algorithm. This is especially true in settings like the one we were working in,
+  where information spans over a limited period of time.
 \end{itemize}
 
-These, however, are fundamental problems of the data we have available. They
+These, however, are fundamental problems of the data we use. They
 affect the results of the algorithm, but do not depend on it so they cannot be
@@
 available and in particular without an external source
-of standard routes, considering this assumption as true is the best
+of standard routes, considering the above assumption as true is the best
 possible approach.
@@
   argument in determining the standard route between two airports. There are two
-  possible approaches to this. The first is to recognize that the informations
+  possible approaches to this. The first is to recognize that the information
   available is not enough, and possibly leave some pairs of airports with
@@
 usually go through 5--10 sectors. Our sectors are smaller --- each flight has
-approximately 20--40 of them --- therefore they represent a much more
+approximately 20--70 of them --- therefore they represent a much more
 fine-grained subdivision of the world map.
@@
 
-A little research showed that an average aircraft procudes about 53 pounds of
+A little research showed that an average aircraft produces about 53 pounds of
 carbon dioxide per mile. We used this value to compute the difference in CO2
@@
 from which we could determine its color. Rather than trying to compute a value
-of ``comsumption'' for each segment, we decided to determine it based on
+of ``consumption'' for each segment, we decided to determine it based on
 how different the route is with respect to the straight-line route. The main
@@
 flights are represented and saved to disk as sequences of sectors, where each
-flight it composed of 20--50 sectors. This gives a very succint representation,
+flight it composed of 20--70 sectors. This gives a very succint representation,
 and we were able to store the output of this phase in just about 4 MB per day.
@@
 area may still be pooly covered, or not covered at all. Even a small poorly
-covered are makes all information on the flights passing through it incomplete,
+covered area makes all information on the flights passing through it incomplete,
 thus unsuitable for our analysis.
@@
 In order to compute the amount of flights diverging from a standard route per
-airline, we needed to associate a an airline to each flight. This is not a
+airline, we needed to associate an airline to each flight. This is not a
 completely trivial task, as airline information is not provided at all by the
@@
 
-We were able to associate an airline to almost all of the flights the resulted
+We were able to associate an airline to almost all of the flights that resulted
 from the route clustering phase. From this information, we then determined how
@@
 
+As a final note, we were also able to conlude that ADS-B data alone is not
+enough to extract information that is sufficiently relevant to other
+applications. The large-scale data sets provided by OpenSky are extremely
+useful, but additional sources of information are necessary if one wants to make
+sense of their raw, unstructured data.
+
 \section{Future work}
@@
 a better distance function.
-A possible alternative way to compute the distance could be to compute the area
+A possible alternative way to estimate the distance could be to compute the area
 delimited by two routes. This area, normalized according to the length of the
-routes, could be a good estimate of how ``different'' two routes are.
+routes, could be a good indication of how ``different'' two routes are.
 A different approach could be to discard the current algorithm and instead use
@@
 
-Currently, the CO2 consumption for the flights is computed in a very simple way,
-by multiplying the distance traveled by the airplanes by an estimate of the
+Currently, the CO2 consumption for the flights is determined in a very simple
+way, by multiplying the distance traveled by the airplanes by an estimate of the
 quantity of CO2 that an average aircraft produces per mile. A better approach
 could use the ICAO number of each single aircraft to index an external dataset
-and discover its characteristics. These properties could be used to compute a
-more precise and realistic estimate of CO2 production.
+containing its characteristics. These properties could be used to compute a
+more precise and realistic estimate of CO2 production for each single aircraft.
 
